Fast and adaptive cointegration based model for forecasting financial time series

### 2017-12-10 slides/defense-slides.tex
before adding second contribution

@@

\section{Contributions}
\begin{frame}
\frametitle{First contribution: Adaptive VECM}
\hspace*{-9mm}
\includegraphics[width=1.15\textwidth]{img/slidingwindow}
\end{frame}

@@

\section{Second Contribution}

\begin{frame}
\frametitle{Second approach}
 Recently, online learning algorithms have been proposed to solve problems with large data sets because of their simplicity and their ability to update the model when new data is
available.
\begin{figure}
\includegraphics[width=0.6\paperwidth]{img/onlinelearning}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Proposal: Online VECM algorithm (OVECM)}
\begin{itemize}
\item Considers only a sliding window of the $L$ most recent data. 
\item Since cointegration vectors represent long-run relationships they vary
little in time. OVECM obtains these vectors only when they need to be updated. 
\item This updating is done when MAPEs of the last $n$ inputs is above a certain
error given by a threshold 
\item OVECM also implements matrix optimisations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Methodology}
OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
sliding window of the most recent data, whereby the models are updated at every
time step. Therefore, we named them SLVECM and SLARIMA.
\end{frame}

\subsection{Parameters}
\begin{frame}
\frametitle{Parameters optimization}
VECM order and ARIMA parameters were selected using AIC.
\begin{table}[ht]
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Execution times}
We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
of the statsmodels library and therefore its execution time is excluded.
%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
%on statsmodels library routine ARIMA.
\begin{center}
\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
           max width=\textwidth}{
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& $L$ & order & e  & Time[s] \\
\hline
OVECM & 100 &p=2  & 0      & 2.492\\
OVECM & 100 &p=2  & 0.0026  & 1.606\\
SLVECM & 100 &p=2& -- & 2.100\\
\hline
OVECM & 400 & p=5  & 0      & 3.513\\
OVECM & 400 &p=5  & 0.0041  & 2.569\\
SLVECM & 400 & p=5 & -- & 3.222\\
\hline
OVECM & 700 &p=3  & 0      & 3.296\\
OVECM & 700 &p=3  & 0.0032  & 2.856\\
SLVECM & 700 &p=3 & -- & 3.581\\
\hline
OVECM & 1000 & p=3 & 0      & 4.387\\
OVECM & 1000 & p=3  & 0.0022  & 2.408\\
SLVECM & 1000 & p=3  & -- & 3.609\\
\hline
\end{tabular}
}\end{center}
\end{frame}

\begin{frame}
\frametitle{Performance accuracy}
\tiny
\begin{table}[H]
\caption{Model measures}
\label{tab:stats}
\begin{center}
\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
\multicolumn{3}{|c|}{Out-of-sample} \\ 
\hline
\hline
Method & $L$ & Parameters &  
MAPE & MAE& RMSE& 
MAPE & MAE& RMSE \\
\hline
 OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
 OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
 OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
 OVECM  &   1000 &  P=3 e=0.0022&  
 \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
 \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
\hline
 SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
 SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
 SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
 SLVECM  &   1000&  P=3 &  
 \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
 \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
\hline
SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}

>>>>

\begin{frame}
\frametitle{Online VECM}
Recently, online learning algorithms have been proposed to solve problems with large data sets main because: 
\begin{itemize}
\item Their simplicity
\item Process one instance at a time
\item The hypothesis is updated every new data arrives
\end{itemize} 
\end{frame}

\begin{frame}
\frametitle{Online learning algorithms}
\begin{algorithm}[H]
\begin{algorithmic}[1]
    \STATE Receives input $\mathbf{x}_t$
    \STATE Makes prediction $\mathbf{\hat{y}}_t$
    \STATE Receives response $\mathbf{y}_t$
    \STATE Incurs loss $l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)$
\end{algorithmic}
\caption{Structure of an Online Learning System}
\end{algorithm}
Performance is later measured after $T$ trials as:
\begin{equation*}
L_T = \sum_{t=1}^T l_t(\mathbf{y}_t,\mathbf{\hat{y}}_t)
\end{equation*}
\end{frame}

@@

\section{Contributions}
\begin{frame}
\frametitle{Adaptive VECM}
\hspace*{-9mm}
\includegraphics[width=1.15\textwidth]{img/slidingwindow}
\end{frame}

@@

%\begin{frame}
%\frametitle{Proposal: Online VECM algorithm (OVECM)}
%\begin{itemize}
%\item Considers only a sliding window of the $L$ most recent data. 
%\item Since cointegration vectors represent long-run relationships they vary
%little in time. OVECM obtains these vectors only when they need to be updated. 
%\item This updating is done when MAPEs of the last $n$ inputs is above a certain
%error given by a threshold 
%\item OVECM also implements matrix optimisations in order to reduce execution time,
%such as updating matrices with new data, removing old data and introducing new
%cointegration vectors.
%\end{itemize}
%\end{frame}
%
%\begin{frame}
%\frametitle{Methodology}
%OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
%sliding window of the most recent data, whereby the models are updated at every
%time step. Therefore, we named them SLVECM and SLARIMA.
%\end{frame}
%
%\subsection{Parameters}
%\begin{frame}
%\frametitle{Parameters optimization}
%VECM order and ARIMA parameters were selected using AIC.
%\begin{table}[ht]
%\label{tab:params}
%\begin{center}
%\begin{tabular}{|c|c|c|}
%\hline
%Windows size $L$ & VECM & ARIMA\\
% $L$ & order $(p)$ & order $(p,d,q)$ \\
%\hline
%100 & 2 & p=2,d=1,q=1\\
%400 & 5 & p=1,d=1,q=1\\
%700 & 3 &p=2,d=1,q=1\\
%1000 &3 & p=2,d=1,q=1\\
%\hline
%\end{tabular}
%\end{center}
%\end{table}
%\end{frame}
%
%\begin{frame}
%\frametitle{Execution times}
%We ran 400 iterations of OVECM and SLVECM . SLARIMA was based on the ARIMA procedure
%of the statsmodels library and therefore its execution time is excluded.
%%because its is not comparable with OVECM and SLVECM. SLARIMA was created based
%%on statsmodels library routine ARIMA.
%\begin{center}
%\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
%           max width=\textwidth}{
%\begin{tabular}{|l|c|c|c|c|c|}
%\hline
%& $L$ & order & e  & Time[s] \\
%\hline
%OVECM & 100 &p=2  & 0      & 2.492\\
%OVECM & 100 &p=2  & 0.0026  & 1.606\\
%SLVECM & 100 &p=2& -- & 2.100\\
%\hline
%OVECM & 400 & p=5  & 0      & 3.513\\
%OVECM & 400 &p=5  & 0.0041  & 2.569\\
%SLVECM & 400 & p=5 & -- & 3.222\\
%\hline
%OVECM & 700 &p=3  & 0      & 3.296\\
%OVECM & 700 &p=3  & 0.0032  & 2.856\\
%SLVECM & 700 &p=3 & -- & 3.581\\
%\hline
%OVECM & 1000 & p=3 & 0      & 4.387\\
%OVECM & 1000 & p=3  & 0.0022  & 2.408\\
%SLVECM & 1000 & p=3  & -- & 3.609\\
%\hline
%\end{tabular}
%}\end{center}
%\end{frame}
%
%\begin{frame}
%\frametitle{Performance accuracy}
%\tiny
%\begin{table}[H]
%\caption{Model measures}
%\label{tab:stats}
%\begin{center}
%\begin{tabular}{|l|l|c|c|c|c|c|c|c|}
%\hline
%\multicolumn{3}{|c|}{Model} & \multicolumn{3}{|c|}{In-sample} &
%\multicolumn{3}{|c|}{Out-of-sample} \\ 
%\hline
%\hline
%Method & $L$ & Parameters &  
%MAPE & MAE& RMSE& 
%MAPE & MAE& RMSE \\
%\hline
% OVECM  &   100  &  P=2 e=0.0026 &  0.00263&  0.00085&  0.00114&  0.00309&  0.00094&  0.00131\\
% OVECM  &   400  &  P=5 e=0.0041&  0.00378&  0.00095&  0.00127&  0.00419&  0.00103&  0.00143\\
% OVECM  &   700  &  P=3 e=0.0032&  0.00323&  0.00099&  0.00130&  0.00322&  0.00097&  0.00132\\
% OVECM  &   1000 &  P=3 e=0.0022&  
% \textbf{0.00175}&  \textbf{0.00062}&  \textbf{0.00087} &  
% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
%\hline
% SLVECM  &   100 &  P=2 &  0.00262&  0.00085&  0.00113&  0.00310&  0.00095&  0.00132\\
% SLVECM  &   400 &  P=5 &  0.00375&  0.00095&  0.00126&  0.00419&  0.00103&  0.00143\\
% SLVECM  &   700 &  P=3 &  0.00324&  0.00099&  0.00130&  0.00322&  0.00098&  0.00132\\
% SLVECM  &   1000&  P=3 &  
% \textbf{0.00174}&  \textbf{0.00061}&  \textbf{0.00087}&  
% \textbf{0.00172}&  \textbf{0.00061}&  \textbf{0.00090}\\
%\hline
%SLARIMA & 100  &p=2 d=1 q=1 & 0.00285  &  0.00110  &  0.00308  &  0.00312  &0.00098  &  0.00144 \\
%SLARIMA & 400  &p=1 d=1 q=1 & 0.00377  &  0.00101  &  0.00128  &  0.00418 & 0.00106 & 0.00145 \\
%SLARIMA & 700  &p=2 d=1 q=1 & 0.00329  &  0.00102  &  0.00136  &  0.00324  & 0.00097  &  0.00133 \\
%SLARIMA & 1000 &p=2 d=1 q=1 & \textbf{0.00281}  & \textbf{0.00074}  &  \textbf{0.00105}  &  \textbf{0.00177} & \textbf{0.00063}  &  \textbf{0.00092} \\
%\hline
%\end{tabular}
%\end{center}
%\end{table}
%\end{frame}
